A sheaf--function dictionary for algebraic tori over local fields

### 2013-08-19 shanghai.tex
Improvements to Part 2.
Maybe it's all there now, mathematically. It certainly needs
re-reading, however,

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}*{sommes trig.} and many people have since referred to such objects
as character sheaves; see for example, \cite{kamgarpour:09a}. 
%\cite{boyarshenko-drinfeld:motivated}*{\S 1.8}.  
Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf--function dictionary to the case of commutative
group schemes which are not necessarily connected or finite type.  The main use
of the term character sheaf is due to Lusztig: it is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in \cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups over algebraically closed fields with cyclic component groups in the series of papers beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

@@

In Section~\ref{sec:GN} we introduce the Greenberg transform of N\'eron models of tori, which produce
group schemes over $\Fq$ from algebraic tori $T$ over $K$.
Section~\ref{sec:bdchar} then uses Theorem~\ref{thm:snake} to give a geometrization of characters of $T$
of bounded depth.
We give a reinterpretation of this result in Section~\ref{sec:quasichar}, using a limit process to simultaneously
geometrize all quasicharacters of $T$.  Sections~\ref{sec:bdchar} and \ref{sec:quasichar} also include
functoriality results that we hope to use in a future discussion of class field theory.
In Section~\ref{sec:transfer} we discuss a transfer principle which allows
a geometric transfer of characters between characteristic $p$ tori to characteristic $0$ tori.
We close the paper with Section~\ref{sec:reciprocity}, in which we suggest that quasicharacter sheaves may play a key role in a geometric and categorical form of the local Langlands correspondence for algebraic tori over non-archimedean local fields.

@@

\begin{proof}
  The identity component $G^0$ of $G$ is a connected algebraic group over $\Fq$, 
  clopen in $G$ \cite{vdGeer-Moonen:AbelianVarieties}*{3.17}. 
  By Proposition~\ref{prop:connected}, every
  quasicharacter sheaf on $G^0$ is isomorphic to a 
  finite quasicharacter sheaf on $G^0$, 
  so to prove the proposition it suffices to show that every 
  finite quasicharacter sheaf on $G^0$ extends to a quasicharacter sheaf on $G$.

@@

For any positive integer $n$, let
\[
%\begin{tikzcd}
\Gr_n : \left( \text{Schemes} \atop \text{lft over $\OK$}\right) \to \left( \text{Schemes} \atop \text{lft over $\Fq$}\right)
%\end{tikzcd}
\]
be the functor produced by precomposing pull-back along $\OK \to \OK/\pK^n$ with the Greenberg functor.
The surjective ring homomorphism $\OK/\pK^{n+1} \to \OK/\pK^n$ determines a natural transformation $P_{n} : \Gr_{n+1} \to \Gr_n$ 
%\Clifton{Do you think I should say more about $P_{n}$?}
\cite{bosch-lutkebohmert-reynaud:NeronModels}*{p. 276}.
For a given Néron model $\TT$ of $T$, set
\[
\GN{T}_n \ceq \Gr_n(\TT)
\qquad \text{and}\qquad
f_n \ceq P_n(\TT).
\]
%
\Clifton{I guess this next bit needs to be re-structured.}
Some fundamental properties of $f_n : \GN{T}_{n+1} \to \GN{T}_n$ are:
 \begin{enumerate}
   \labitem{(GN.0)}{points} $\GN{T}_n(\Fq) = \TT(\OK/\pK^n)$, canonically.
  \labitem{(GN.1)}{smooth} $\GN{T}_n$ is a smooth commutative group scheme
  locally of finite type over $\Fq$ and the component group of $\GN{T}_n$ is
  geometrically finitely generated: $\underline{T}$ is a smooth group scheme over $\OK$, by construction; since smoothness is preserved under base change, $\TT_n \ceq \underline{T} \times_\Spec{\OK} \Spec{\OK/\pK^n}$ is a smooth group scheme over an Artinian ring with perfect residue field; the Greenberg functor preserves smoothness, so $\GN{T}_n$ is a smooth group scheme over $\Fq$; the identity component of any smooth group scheme is geometrically connected, hence connected, and the Greenberg functor takes the geometric identity component of $\TT_n$ to the geometric identity component of $\GN{T}_n$;
  in fact, $\pi_0(\GN{T}_n) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$;
  this group is finitely generated by \cite{xarles:93a}*{2.18}.
  \labitem{(GN.2)}{truncation}
  The morphism $f_n : \GN{T}_{n+1} \to \GN{T}_n$ is a surjective morphism of group schemes. To see this, it suffices to find a morphism of schemes $i_n : \GN{T}_{n}\hookrightarrow \GN{T}_{n+1}$ such that $f_n\circ i_n = \id_{\GN{T}_n}$. Base change along $\OK/\pK^{n+1} \to \OK/\pK^n$ determines a closed immersion of schemes $\TT_{n+1} \hookrightarrow \TT_n$ (not a morphism of group schemes). Applying the Greenberg functor produces a closed immersion of schemes $i_n : \GN{T}_{n} \hookrightarrow \GN{T}_{n+1}$. 
\Clifton{I'm working on this.}
Although $i_n$ is not a morphism of group schemes, it is a section of $f_n : \GN{T}_{n+1}\to \GN{T}_n$.
  \end{enumerate}
%
We remark that $f_n : \GN{T}_{n+1} \to \GN{T}_n$ is not induced from a morphism of Néron models.

@@

For any $n< m$ (positive integers) define $f_{n\leq m} : \GN{T}_{m} \to \GN{T}_{n}$ by $f_{n\leq m}\ceq f_n \circ f_{n+1} \circ \cdots \circ f_{m-1}$; also set $f_{n\leq n} \ceq \id_{\GN{T}_n}$.
%\Clifton{Now I wonder if $f_{n\leq m}$ is a discrete isogeny!!!  What's the covering group of $f_{n\leq m}$?  Must be the finite group $\GN{T}_{m}(\Fq)/\GN{T}_n(\Fq)$, which could be $\TT^0(\pK^n/\pK^{m})$. We might be able to build some very special finite quasicharacter sheaves on $\GN{T}_n$ this way.}
Let $f_{n\leq m}^* : \QC(\GN{T}_n) \to \QC(\GN{T}_{m})$ be the functor promised by Proposition~\ref{prop:pullback}.
%
%\begin{proposition}\label{prop:nm}
The functor $f_{n\leq m}^* : \QC(\GN{T}_n) \to \QC(\GN{T}_{m})$ is full and faithful, by \ref{truncation}.
%\end{proposition}
%
%\begin{proof}
%
%\end{proof}

We now define a category which allows us to extend the notion of quasicharacter sheaf to the pro-group scheme $\GN{T}$.

>>>>

We want to situate the choice of term ``quasicharacter sheaf'' in a historical context.
Analogous sheaves on connected, commutative algebraic groups were first studied
by Deligne \cite{deligne:SGA4.5}*{sommes trig.} and many people have since referred to such objects
as character sheaves; see for example, \cite{kamgarpour:09a}. 
%\cite{boyarshenko-drinfeld:motivated}*{\S 1.8}.  
Our quasicharacter sheaves evolved from this notion, and Theorem~\ref{thm:snake}
is a direct generalization of the sheaf--function dictionary to the case of commutative
group schemes which are not necessarily connected or finite type.  The main use
of the term character sheaf is due to Lusztig: it is applied to certain perverse sheaves on connected reductive algebraic groups over algebraically closed fields in \cite{lusztig:85a}*{Def.~2.10} and to certain perverse sheaves on reductive groups over algebraically closed fields with cyclic component groups in the series of papers beginning with \cite{lusztig:disconnected1}.  
When commutative, such groups are extensions of $\ZZ/n\ZZ$ by a torus,
in which case it is not difficult to relate Frobenius-stable character sheaves to our quasicharacter sheaves.

@@

In Section~\ref{sec:GN} we review the Greenberg transform of N\'eron models of tori, which produces
group schemes over $\Fq$ from algebraic tori $T$ over $K$.
Section~\ref{sec:bdchar} then uses Theorem~\ref{thm:snake} to give a geometrization of characters of $T$
of bounded depth.
We give a reinterpretation of this result in Section~\ref{sec:quasichar}, using a limit process to simultaneously
geometrize all quasicharacters of $T$.  Sections~\ref{sec:bdchar} and \ref{sec:quasichar} also include
functoriality results that we hope to use in a future discussion of class field theory.
In Section~\ref{sec:transfer} we discuss a transfer principle which allows
a geometric transfer of characters between characteristic $p$ tori and characteristic $0$ tori.
We close the paper with Section~\ref{sec:reciprocity}, in which we suggest that quasicharacter sheaves may play a key role in a geometric and categorical form of the local Langlands correspondence for algebraic tori over non-archimedean local fields.

@@

\begin{proof}
  The identity component $G^0$ of $G$ is a connected algebraic group over $\Fq$, 
  \Clifton{This needs to be treated more carefully!}
  clopen in $G$ \cite{vdGeer-Moonen:AbelianVarieties}*{3.17}. 
    By Proposition~\ref{prop:connected}, every
  quasicharacter sheaf on $G^0$ is isomorphic to a 
  finite quasicharacter sheaf on $G^0$, 
  so to prove the proposition it suffices to show that every 
  finite quasicharacter sheaf on $G^0$ extends to a quasicharacter sheaf on $G$.

@@

For any positive integer $n$, let
\[
%\begin{tikzcd}
\Gr_n : \left( \text{Schemes} \atop \text{lft over $\OK$}\right) \to \left( \text{Schemes} \atop \text{lft over $\Fq$}\right)
%\end{tikzcd}
\]
be the functor produced by precomposing pull-back along $\OK \to \OK/\pK^n$ with the Greenberg functor.
The surjective ring homomorphism $\OK/\pK^{n+1} \to \OK/\pK^n$ determines a natural transformation $P_{n} : \Gr_{n+1} \to \Gr_n$ 
%\Clifton{Do you think I should say more about $P_{n}$?}
\cite{bosch-lutkebohmert-reynaud:NeronModels}*{p. 276}.
For a given Néron model $\TT$ of $T$, set
\[
\GN{T}_n \ceq \Gr_n(\TT)
\qquad \text{and}\qquad
f_n \ceq P_n(\TT).
\]
Some fundamental properties of $f_n : \GN{T}_{n+1} \to \GN{T}_n$ are:
 \begin{enumerate}
   \labitem{(GN.0)}{points} $\GN{T}_n(\Fq) = \TT(\OK/\pK^n)$, canonically.
  \labitem{(GN.1)}{smooth} 
  $\GN{T}_n$ is a smooth commutative group scheme locally of finite type over $\Fq$ 
  and the component group of $\GN{T}_n$ is geometrically finitely generated. 
  Note that $\underline{T}$ is a smooth group scheme over $\OK$, by construction. 
  Since smoothness is preserved under base change, 
  $\TT_n \ceq \underline{T} \times_\Spec{\OK} \Spec{\OK/\pK^n}$ 
  is a smooth group scheme over an Artinian ring with perfect residue field. 
  The Greenberg functor preserves smoothness, 
  so $\GN{T}_n$ is a smooth group scheme locally of finite type over $\Fq$. 
  The Greenberg functor takes the geometric identity component of $\TT_n$ 
  to the geometric identity component of $\GN{T}_n$;
  in fact, $\pi_0(\GN{T}_n) = \pi_0(\TT) \times_{\Spec{\OK}} \Spec{\Fq}$ and
  this group is finitely generated by \cite{xarles:93a}*{2.18}.
  \labitem{(GN.2)}{truncation}
  The morphism $f_n : \GN{T}_{n+1} \to \GN{T}_n$ is a morphism of group schemes over $\Fq$. 
  This follows from the fact that $P_{n}$ is a natural transformation between additive functors.
  We remark that $f_n : \GN{T}_{n+1} \to \GN{T}_n$ is not induced from a morphism of Néron models.
\end{enumerate}

@@

For any $n< m$ (positive integers) define $f_{n\leq m} : \GN{T}_{m} \to \GN{T}_{n}$ by $f_{n\leq m}\ceq f_n \circ f_{n+1} \circ \cdots \circ f_{m-1}$; also set $f_{n\leq n} \ceq \id_{\GN{T}_n}$.
Note that $f_{n\leq m}:  \GN{T}_{m} \to \GN{T}_{n}$ is a morphism of group schemes over $\Fq$.
Let $f_{n\leq m}^* : \QC(\GN{T}_n) \to \QC(\GN{T}_{m})$ be the functor produced by applying Proposition~\ref{prop:pullback} to $f_{n\leq m}$.
We now define a category which allows us to extend the notion of quasicharacter sheaf to the pro-group scheme $\GN{T}$.
